Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-10-17 prop_letter.tex
prop_letter.tex \section{Gleason} Updating...

\section{Born\label{sec:Born}}

\yutsung{Notice that the use of quantum is minimized in this section.
In another word, the results of this section should apply to both
quantum and classical IVPMs.}

Recall in Sec.~\ref{sec:fuzzy}, we illustrate the non-constructive
postulates for the quantum probability measures first, and then introduce
the Born rule which induces a quantum probability measures from each
state~\citep{Born1983,peres1995quantum,544199,Jaeger2007}. Since
the postulates for QIVPMs in Sec.~\ref{sec:Interval-Uncertainty}
are also non-constructive, we want to find a simple rule to construct
a QIVPM as well.

However, the composited function~$\iota\circ\mu$ may or may not
be a QIVPM. For example, in a four-dimensional Hilbert space, consider
the function 
\begin{equation}
\iota_{1}\left(x\right)=\begin{cases}
\imposs\,, & \textrm{ if }x=0\,;\\
\left[0,\tfrac{1}{2}\right]\,, & \textrm{ if }x\in\left(0,\tfrac{1}{4}\right]\,;\\
\unknown\,, & \textrm{ if }x\in\left(\tfrac{1}{4},\tfrac{3}{4}\right)\,;\\
\left[\tfrac{1}{2},1\right]\,, & \textrm{ if }x\in\left[\tfrac{3}{4},1\right)\,;\\
\necess\,, & \textrm{ if }x=1\,.
\end{cases}\label{eq:iota}
\end{equation}
While the function~$\iota_{1}\circ\mu_{\rho_{0}}^{\mathrm{\mathrm{B}}}$
induced by $\rho_{0}=\frac{1}{4}\cdot\proj{0}+\frac{1}{4}\cdot\proj{1}+\frac{1}{4}\cdot\proj{2}+\frac{1}{4}\cdot\proj{3}$
is a QIVPM, the function~$\iota_{1}\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$
induced by $\rho_{1}=\frac{1}{8}\cdot\proj{0}+\frac{1}{8}\cdot\proj{1}+\frac{1}{8}\cdot\proj{2}+\frac{5}{8}\cdot\proj{3}$
is \emph{not} a QIVPM~\citep{HOSTunpublished}. In order not to verify
whether $\iota\circ\mu$ is a QIVPM for each individual case, it is
desirable to have some criteria to discriminate different cases. We
will start from the simpler criteria for the dimension~$d\le3$,
and then the general case.

\begin{thm}\label{thm:QRVPM2QIVPM-3}In  for any quantum probability
measure~$\mu:\events\rightarrow\left[0,1\right]$.\end{thm}

The situation for $d=2$ is trivial. On a three-dimensional Hilbert
space, to prove that $\iota\circ\mu$ is a QIVPM, it is sufficient
to check Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
\begin{equation}
\begin{aligned} & \iota\left(\mu\left(P_{0}+P_{1}-P_{0}P_{1}\right)\right)+\iota\left(\mu\left(P_{0}P_{1}\right)\right)\\
\subseteq{} & \iota\left(\mu\left(P_{0}\right)\right)+\iota\left(\mu\left(P_{1}\right)\right)\,.
\end{aligned}
\label{eq:QuantumInterval-valuedProbability-Inclusion-case}
\end{equation}
 Depending on the relation between $P_{0}$ and $P_{1}$, we have
three possibilities:
\begin{itemize}
\item $P_{0}$ and $P_{1}$ are orthogonal.
\item Their complements~$\mathbb{1}-P_{0}$ and $\mathbb{1}-P_{1}$ are
orthogonal.
\item One of them projects onto a subspace of what the other one projects
onto.
\end{itemize}
\noindent All cases deduce Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion-case})
easily~\citep{HOSTunpublished} so that $\iota\circ\mu$ is indeed
a QIVPM.

On the one hand, $\iota_{1}$ defined in Eq.~(\ref{eq:iota}) satisfies
Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add})~\citep{HOSTunpublished}
so that every function~$\iota_{1}\circ\mu$ on a three-dimensional
Hilbert space is a QIVPM. On the other hand, $\iota_{1}\circ\mu_{\rho_{1}}^{\mathrm{\mathrm{B}}}$
is not a QIVPM on four-dimensional Hilbert space. Hence, to guarantee
$\iota\circ\mu$ is always a QIVPM, we need to strengthen Eq.~(\ref{eq:iota-add})
and consider the following conditions.

\yutsung{Another idea might be fixing a quantum probability measure~$\mu$,
whether there is a length of intervals, such that if $\iota$ maps
to intervals tighter than the given length, then $\iota\circ\mu$
is always a QIVPM... I don't know the answer of this question, and
it might not be very easy...}

\yutsung{The other idea: We never proved a non-convex QIVPM always
have a non-empty core or find an counterexample that a non-convex
QIVPM has empty core. Moreover, we have no idea the relation between
its expectation value and the Choquet integral... Depending the on
situation, the convex condition may not be necessary, but this might
be hard to prove...}

@@

In other words, if the measurement intervals are too imprecise and
the QIVPM are defined so wildly, the results we get might contradict
to each other, and exclude any possible state from being consistent
with the measurement results. To address this problem, we are going
to consider a simple rule which can induce well-behaved QIVPMs by
quantum probability measures systematically. Recall when we generalized
quantum probability measures to QIVPMs, we changed the range from
the unit interval~$\left[0,1\right]$ to a set of intervals~$\mathscr{I}$.
This replacement suggests us to find a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
which can be used to replace a real-valued probability~$\mu\left(P\right)$
by an interval-valued one~$\iota\left(\mu\left(P\right)\right)$.
In other words, compositing $\iota$ with a quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
produces an interval-valued function~$\iota\circ\mu:\events\overset{\mu}{\rightarrow}\left[0,1\right]\overset{\iota}{\rightarrow}\mathscr{I}$
which has the same domain and range as a QIVPM. Although there are
infinitely many different choice of $\iota$, the function~$\iota$
satisfies the following homomorphism-like properties are of particular
interest, because it guarantee $\iota\circ\mu$ is always a QIVPM.

\begin{definition}Given a collection of intervals $\mathscr{I}$
including $\imposs$ and $\necess$, a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
is called a \emph{morphism}\yutsung{Haven't found a better name so
far, but haven't proved that it is a morphism in the sense of category
theory as well...}if $\iota:\left[0,1\right]\rightarrow\mathscr{I}$
is subject to the following constraints:\begin{subequations}\label{eq:iota-constraints}
\begin{eqnarray}
 &  & \iota(0)=\imposs\,,\\
 &  & \iota(1)=\necess\,,\\
 &  & \iota\left(1-x\right)=\necess-\iota\left(x\right)\,,
\end{eqnarray}
\end{subequations}and satisfies that for any three numbers~$x_{0}$,
$x_{1}$, and $x_{2}\in\left[0,1\right]$, their sum~$x_{0}+x_{1}+x_{2}\in\left[0,1\right]$
implies
\begin{equation}
\iota\left(x_{0}+x_{1}+x_{2}\right)+\iota\left(x_{2}\right)\subseteq\iota\left(x_{0}+x_{2}\right)+\iota\left(x_{1}+x_{2}\right)\,.\label{eq:iota-Inclusion}
\end{equation}
\end{definition}

\noindent Then, $\iota$ is a morphism precisely characterizes that
$\iota\circ\mu$ is always a IVPM no matter in classical or quantum
sense.

@@

Since the similarity between classical and quantum cases, we will
only prove the equivalence between Statement~\ref{enu:iota-subject-to}
and \ref{enu:iota-mu-QIVPM}. It is clear that $\iota$ is a morphism
implies $\iota\circ\mu$ is always a QIVPM, and the rest of the proof
is to prove that if $\iota$ is not a morphism, then $\iota\circ\mu$
might not be a QIVPM. If $\iota$ doesn't satisfy Eq.~(\ref{eq:iota-Inclusion}),
we can find three particular numbers~$x_{0}$, $x_{1}$, and $x_{2}\in\left[0,1\right]$
such that $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$, but $\iota\left(x_{0}+x_{1}+x_{2}\right)+\iota\left(x_{2}\right)\nsubseteq\iota\left(x_{0}+x_{2}\right)+\iota\left(x_{1}+x_{2}\right)$.
Consider the state $\rho=x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}+\left(1-x_{0}-x_{1}-x_{2}\right)\proj{3}$.
Then, the function~$\iota\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}:\events\rightarrow\mathscr{I}$
fails to satisfy Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
\begin{eqnarray*}
 &  & \iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{0}+\proj{1}+\proj{2}\right)\right)+\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{2}\right)\right)\\
 & \nsubseteq & \iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{0}+\proj{2}\right)\right)+\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{1}+\proj{2}\right)\right)\,,
\end{eqnarray*}
and cannot be a QIVPM. Therefore, $\iota$ is a morphism if and only
if $\iota\circ\mu$ is always a QIVPM.

Although the composited function~$\iota\circ\mu$ may not guarantee
to be a QIVPM when $\iota$ is not a morphism, it may still be a QIVPM
under so suitable conditions. For example, given a Hilbert space of
dimension~$d\le3$, we can guarantee $\iota\circ\mu$ is a QIVPM
when $\iota$ satisfies Eqs.~(\ref{eq:iota-constraints}) and a weaker
condition: for any numbers~$x_{0}$ and $x_{1}\in\left[0,1\right]$,
their sum~$x_{0}+x_{1}\in\left[0,1\right]$ implies
\begin{equation}
\iota\left(x_{0}+x_{1}\right)\subseteq\iota\left(x_{0}\right)+\iota\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}

In contrast with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM},
if a QIVPM~$\iota\circ\mu_{\rho}^{\mathrm{B}}$ is induced by a conventional
quantum probability measure~$\mu_{\rho}^{\mathrm{B}}$ and a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
, there is always a state consistent with it because $\rho\in\coreBorn\left(\iota\circ\mu_{\rho}^{\mathrm{B}},\events\right)$.
Moreover, the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower. To state this result, we introduce a few technical notions.
First, the norm of a pure state~$\ket{\psi}$ is defined as usual
$||\psi||=\sqrt{\ip{\psi}{\psi}}$. Then, the norm of the difference
of two states~$\rho$ and $\rho'$ is just the usual operator norm
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert \left(\rho-\rho'\right)\ket{\psi}\right\Vert \,,
\end{equation}
which is also commonly known as $2$-norm or spectral norm~\citep{RobertsVarberg1973,peres1995quantum,GolubVanLoan1996,Foucart2012}.

@@

\begin{thm}\label{thm:convex-uncountable}Let 
\begin{equation}
\mathscr{I}_{0}=\left\{ \necess,\imposs,\unknown\right\} \,.
\end{equation}
For any function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
described in Thm.~\ref{thm:iota-statements}, either $\mathscr{I}=\mathscr{I}_{0}$
or $\mathscr{I}$ contains uncountably many intervals.\end{thm}

In order to find a function~$\iota$ satisfying Eq.~(\ref{eq:iota-Inclusion}),
we decompose it into two functions: its left-end and right-end~$\left[\iotal\left(x\right),\iotar\left(x\right)\right]=\iota\left(x\right)$.
By Eq.~(\ref{eq:iota-Inclusion}), the left-end function~$\iotal:\left[0,1\right]\rightarrow\left[0,1\right]$
is Wright-convex, i.e., $\iotal\left(x_{0}+x_{1}+x_{2}\right)+\iotal\left(x_{2}\right)\ge\iotal\left(x_{0}+x_{2}\right)+\iotal\left(x_{1}+x_{2}\right)$~\citep{Wright1954,RobertsVarberg1973,PecaricTong1992}.
It is well-known that a Wright-convex function is midpoint-convex
or Jensen-convex~\citep{Klee_1956,Kenyon_1956,RobertsVarberg1973,PecaricTong1992},
i.e., we have
\begin{equation}
\frac{\iotal\left(x_{0}\right)+\iotal\left(x_{1}\right)}{2}\ge\iotal\left(\frac{x_{0}+x_{1}}{2}\right)\,.
\end{equation}
Together with the fact that $\iotal$ maps to a bounded interval~$\left[0,1\right]$,
the left-end function~$\iotal$ must be continuous on the unit open
interval~$\left(0,1\right)$~\citep{Jensen_1906,RobertsVarberg1973}.
Therefore, either $\iota$ maps every number to the same interval,
or the number of intervals that $\iota$ maps to must be uncountable.

>>>>

@@

In other words, if the measurement intervals are too imprecise and
the QIVPM are defined so wildly, the results we get might contradict
to each other, and exclude any possible state from being consistent
with the measurement results. To address this problem, we are going
to consider a simple rule which can induce well-behaved QIVPMs by
quantum probability measures systematically. Recall when we generalized
quantum probability measures to QIVPMs, we changed the range from
the unit interval~$\left[0,1\right]$ to a set of intervals~$\mathscr{I}$.
This replacement suggests us to find a function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
which can be used to replace a real-valued probability~$\mu\left(P\right)$
by an interval-valued one~$\iota\left(\mu\left(P\right)\right)$.
In other words, compositing $\iota$ with a quantum probability measure~$\mu:\events\rightarrow\left[0,1\right]$
produces an interval-valued function~$\iota\circ\mu:\events\overset{\mu}{\rightarrow}\left[0,1\right]\overset{\iota}{\rightarrow}\mathscr{I}$
which has the same domain and range as a QIVPM. In order to guarantee
$\iota\circ\mu$ is always a QIVPM, we consider the following homomorphism-like
properties for $\iota$.

\begin{definition}[{morphism}]{\label{def:morphism}}Given a collection
of intervals $\mathscr{I}$ including $\imposs$ and $\necess$, a
function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$ is called
a \emph{morphism}\yutsung{We can give it a better name. If sticking
with this name, we may need to prove that it is a morphism in the
sense of category theory or something...}if $\iota:\left[0,1\right]\rightarrow\mathscr{I}$
is subject to the following constraints:\begin{subequations}\label{eq:iota-constraints}
\begin{eqnarray}
 &  & \iota(0)=\imposs\,,\\
 &  & \iota(1)=\necess\,,\\
 &  & \iota\left(1-x\right)=\necess-\iota\left(x\right)\,,
\end{eqnarray}
\end{subequations}and satisfies that for any three numbers~$x_{0}$,
$x_{1}$, and $x_{2}\in\left[0,1\right]$, their sum~$x_{0}+x_{1}+x_{2}\in\left[0,1\right]$
implies
\begin{equation}
\iota\left(x_{0}+x_{1}+x_{2}\right)+\iota\left(x_{2}\right)\subseteq\iota\left(x_{0}+x_{2}\right)+\iota\left(x_{1}+x_{2}\right)\,.\label{eq:iota-Inclusion}
\end{equation}
\end{definition}

\noindent Because the similarity between Def.~\ref{def:morphism}
and Def.~\ref{def:QIVPM}, $\iota\circ\mu$ should always be a IVPM
no matter in classical or quantum case. Surprisingly, if $\iota\circ\mu$
is always a IVPM, then $\iota$ must be a morphism.

@@

As we have discussed, Statement~\ref{enu:iota-subject-to} implies
\ref{enu:iota-mu-QIVPM}. The rest of the proof is to prove that if
$\iota$ is not a morphism, then $\iota\circ\mu$ might not be a QIVPM
for any quantum probability measure~$\mu$. If $\iota$ doesn't satisfy
Eq.~(\ref{eq:iota-Inclusion}), we can find three particular numbers~$x_{0}$,
$x_{1}$, and $x_{2}\in\left[0,1\right]$ such that $x_{0}+x_{1}+x_{2}\in\left[0,1\right]$,
but $\iota\left(x_{0}+x_{1}+x_{2}\right)+\iota\left(x_{2}\right)\nsubseteq\iota\left(x_{0}+x_{2}\right)+\iota\left(x_{1}+x_{2}\right)$.
Consider the state $\rho=x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}+\left(1-x_{0}-x_{1}-x_{2}\right)\proj{3}$.
Then, the function~$\iota\circ\mu_{\rho}^{\mathrm{\mathrm{B}}}:\events\rightarrow\mathscr{I}$
fails to satisfy Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
\begin{eqnarray*}
 &  & \iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{0}+\proj{1}+\proj{2}\right)\right)+\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{2}\right)\right)\\
 & \nsubseteq & \iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{0}+\proj{2}\right)\right)+\iota\left(\mu_{\rho}^{\mathrm{\mathrm{B}}}\left(\proj{1}+\proj{2}\right)\right)\,,
\end{eqnarray*}
and cannot be a QIVPM. Therefore, $\iota$ is a morphism if and only
if $\iota\circ\mu$ is always a QIVPM. Similarly, Statement~\ref{enu:iota-subject-to}
and \ref{enu:iota-mu-CIVPM} are equivalent as well~\citep{HOSTunpublished}.

Even if $\iota$ is not a morphism, the composited function~$\iota\circ\mu$
may still be a QIVPM under so suitable conditions. For example, Eq.~(\ref{eq:iota-Inclusion})
can be weaken as follow: for any numbers~$x_{0}$ and $x_{1}\in\left[0,1\right]$,
their sum~$x_{0}+x_{1}\in\left[0,1\right]$ implies
\begin{equation}
\iota\left(x_{0}+x_{1}\right)\subseteq\iota\left(x_{0}\right)+\iota\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
Then, given a Hilbert space of dimension~$d\le3$, when $\iota$
satisfies Eqs.~(\ref{eq:iota-constraints}) and Eq.~(\ref{eq:iota-add}),$\iota\circ\mu$
still must be a QIVPM~\citep{HOSTunpublished}.

In contrast with the QIVPM defined in Table~\ref{tab:non-Born-QIVPM},
there is always a state consistent with the composited QIVPM~$\iota\circ\mu_{\rho}^{\mathrm{B}}$~\citep{HOSTunpublished}.
Moreover, the collection of states consistent with $\iota\circ\mu_{\rho}^{\mathrm{B}}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower. To state this result, we introduce a few technical notions.
First, the norm of a pure state~$\ket{\psi}$ is defined as usual
$\left\Vert \psi\right\Vert =\sqrt{\ip{\psi}{\psi}}$. Then, the norm
of the difference of two states~$\rho$ and $\rho'$ is just the
usual operator norm
\begin{equation}
\left\Vert \rho-\rho'\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert \left(\rho-\rho'\right)\ket{\psi}\right\Vert \,,
\end{equation}
which is also commonly known as $2$-norm or spectral norm~\citep{RobertsVarberg1973,peres1995quantum,GolubVanLoan1996,Foucart2012}.

@@

Since morphisms have many desirable properties, the intervals it mapping
to either is trivial or contains infinity intervals.

\begin{thm}\label{thm:convex-uncountable}For any function~$\iota:\left[0,1\right]\rightarrow\mathscr{I}$
described in Thm.~\ref{thm:iota-statements}, either $\mathscr{I}=\mathscr{I}_{0}$
defined in Eq.(\ref{eq:3-value-intervals}) or $\mathscr{I}$ contains
uncountably many intervals.\end{thm}

Since $\iota$ maps to intervals, we can decompose it into two functions:
its left-end and right-end~$\left[\iotal\left(x\right),\iotar\left(x\right)\right]=\iota\left(x\right)$.
By Eq.~(\ref{eq:iota-Inclusion}), the left-end function~$\iotal:\left[0,1\right]\rightarrow\left[0,1\right]$
is Wright-convex, i.e., $\iotal\left(x_{0}+x_{1}+x_{2}\right)+\iotal\left(x_{2}\right)\ge\iotal\left(x_{0}+x_{2}\right)+\iotal\left(x_{1}+x_{2}\right)$~\citep{Wright1954,RobertsVarberg1973,PecaricTong1992}.
It is well-known that a Wright-convex function is midpoint-convex
or Jensen-convex~\citep{Klee_1956,Kenyon_1956,RobertsVarberg1973,PecaricTong1992},
i.e., we have
\begin{equation}
\frac{\iotal\left(x_{0}\right)+\iotal\left(x_{1}\right)}{2}\ge\iotal\left(\frac{x_{0}+x_{1}}{2}\right)\,.
\end{equation}
Together with the fact that $\iotal$ maps to a bounded interval~$\left[0,1\right]$,
the left-end function~$\iotal$ must be continuous on the unit open
interval~$\left(0,1\right)$~\citep{Jensen_1906,RobertsVarberg1973}.
Therefore, either $\iota$ maps every number to the same interval,
or the number of intervals that $\iota$ maps to must be uncountable.
